---
title: "A brief introduction to mplot"
author: "Garth Tarr"
date: "`r Sys.Date()`"
output: html_document
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{A brief introduction to mplot}
-->

# A brief introduction to mplot

The **mplot** package implements a range of model selection procedures and model stability plots designed to provide users with the information they need to make informed decisions about model selection issues in linear and generalised linear models.  Future versions of this package will also consider mixed models.

There are two main componenets: model selection via the adaptive fence procedure (Jiang et. al., 2009) and model stability curves as described in Müller and Welsh (2010).

### Example 1

The adaptive fence procedure is implemented through the function `af` as follows:

```{r artificial.eg, tidy=FALSE}
require(mplot,quietly=TRUE)
op = options(gvis.plot.tag = "chart")
n = 100
set.seed(11)
e = rnorm(n)
x0 = 1
x1 = rnorm(n)
x2 = rnorm(n)
x3 = x1^2
x4 = x2^2
x5 = x1*x2
y = x0 + x1 + x2 + e
dat = data.frame(y,x1,x2,x3,x4,x5)
lm1 = lm(y~., data=dat)
af1 = af(lm1, n.cores=4)
summary(af1)
```

```{r af1plot, results='asis', tidy=FALSE}
plot(af1)
```

In this example, we can see that the first (and only) peak in the plot of $p^*$ against $c$ occurs for the model of $y$ against $x_1$ and $x_2$.  The $c^*$ value is `r round(af1$bestOnly$c.star,1)`, and the model selected as a result of using that $c^*$ value is given in the summary output.  

Note that the range of $c$ can be set manually using the `c.max` argument or it can be selected using an initial application of forward and backward searches.

```{r af.stepwise, tidy=FALSE}
af1.step = af(lm1, n.cores=4, initial.stepwise=TRUE)
summary(af1.step)
```

```{r af1.step.plot, results='asis', tidy=FALSE}
plot(af1.step)
```

### Example 2


If you have something like
```{r afeg2}
n = 100
set.seed(11)
e = rnorm(n)
x0 = 1
x1 = rnorm(n)
x2 = rnorm(n)
x6 = rep(c("A","B","C","D"),n/4)
x3 = x1^2
x4 = x2^2
x5 = x1*x2
gp = as.numeric(factor(x6)) 
yI = x0 + x1 + x2 + gp + e
datI = data.frame(yI,x1,x2,x3,x4,x5,x6)
lmI = lm(yI~., data=datI)
af2 = af(lmI, n.cores=4, initial.stepwise=TRUE)
summary(af2)
```

The model chosen by the automated procedure is wrong, however, if we look at the plot, we can see that there was an earlier peak, though in this instance, it didn't achieve a level higher than the second peak, so it wasn't picked.

```{r af2plot, results='asis', tidy=FALSE}
plot(af2)
```

Take home message: it is important to look at the plot!

### Example 3

It works basically the same for generalised linear models, except a LOT slower as we are unable to take advantage of the wonders of the **leaps** package.  Instead we call on the **bestglm** package which essentially does an exhaustive search when the link function is not normal.  The following quite simple example takes a few minutes to run.

```{r binom.eg, results='hide', tidy=FALSE, eval=FALSE}
require(mvtnorm)
n=200
set.seed(11)
p = 4
cor.mat = diag(p)
cor.mat[1,2] = cor.mat[2,1] = 0.7
X = rmvnorm(n,sigma=cor.mat)
beta = c(1,2.1,0,0.9,1.1)
z = cbind(1,X)%*%beta
pr = 1/(1+exp(-z))
y = rbinom(n,1,pr) 
glm.df = data.frame(y,X)
glm1 = glm(y~.,data=glm.df,family="binomial")
af.glm = af(glm1, n.cores=4, c.max=14, n.c=40, B = 100)
summary(af.glm)
plot(af.glm)
```


### Real world example

If there exists a *true* model, the adaptive fence tends to find it quite easily.  In practice, there is rarely such a well defined and readily identifiable data generating process and it is the role of the investigator to make a judgement call as to which model *best* describes the underlying process.  A nice feature of the adaptive fence procedure is its ability to guide the investigator and highlight plausible candidate models.

Consider the body fat example of Johnson (1996).  

```{r bdat.full.plot, results='asis', tidy=FALSE}
data(bodyfat)
bfat.lm = lm(Bodyfat~.,data=subset(bodyfat,select=-Id))
bfat.full = af(bfat.lm, n.cores=4, initial.stepwise=FALSE)
plot(bfat.full)
```

Note that if we consider an extended range of possible values for the parameter $c$, by setting `initial.stepwise=FALSE`, it appears that there is one dominant variable, `abdo`, and hence one might chose a simple linear regression of `Bodyfat` on `abdo`.

Instead, one might like to incorporate some additional information to better inform our choice.  In particular, let us consider the number of variables that AIC and BIC forward and backward stepwise procedures and investigate only models of similar size.

```{r bfat.stepwise.plot, results='asis', tidy=FALSE}
bfat.rest = af(bfat.lm, n.cores=4, n.c=50, c.max = 20)
plot(bfat.rest)
```

Doing this, we focus attention on smaller $c$ values, i.e. larger model sizes.  We can see that the regression of `Bodyfat` on `weight` and `abdo` is chosen over a large range of c values.  However, it is also important to note that in this range, the pstar values are all quite low, suggesting that no second variable (in addition to `abdo`) is dominant.

```{r}
summary(bfat.rest)
```


## Model selection curves

The standard plots are the model stability and variable inclusion plots.  Both of these are performed using the `vis()` function.  We illustrate using the linear model defined previously:

```{r vis1.eg, results='asis', tidy=FALSE}
vis1 = vis(lmI,n.cores=4,B=50)
```

### Model stability plots

There are two variants here.  The first simply iterates over all possible models at each model size and returns a scatter plot showing a goodness of fit measure (-2*loglik) against model size.  You need to specify which variable to highlight (this is done best via the Shiny interface, see below).

```{r msc, results='asis', tidy=FALSE}
plot(vis1,which="lvk",highlight="x1")
```

The second, performs a simple weighted bootstrap to see which models perform *best* at conditional on the model size.

```{r boot, results='asis', tidy=FALSE}
plot(vis1,which="boot",highlight="x1")
```

### Variable inclusion plots

The variable inclusion plots use that same bootstrap simulations as the second model selection curve plot and identifies how often each variable is selected as being in the *best* model for a given penalty.

```{r vip, results='asis', tidy=FALSE}
plot(vis1,which="vip")
```

### Bodyfat example

```{r vis, results='hide', tidy=FALSE}
data(bodyfat)
vis.bf = vis(bfat.lm, n.cores=4, nvmax = 4)
```

```{r, lvk.bf, results='asis'}
plot(vis.bf, which="lvk",highlight = "Abdo")
```

```{r, boot2, results='asis'}
plot(vis.bf, which="boot", highlight = "Abdo")
```

```{r, vip2, results='asis'}
plot(vis.bf, which="vip")
```


## Shiny interface

All of these can be accessed, and interacted with, via a Shiny web interface.  The first tab is a scatterplot of the data, you start by selecting some variables and go from there - nothing technical here, just a nice interactive way to view a moderate dimensional data set.  You can select points and see them highlighted across all plots and (optionally) see a data table.

The other tabs show the model stability plots, the fence plot and the variable inclusion plots.

This functionality is currently acessed through the `mplot()` function. For the above example, you'd use the following code:

`mplot(lm1, vis=vis1, fence=af1)`